\large \textbf{Title:} Computational models reveal how cleaner fish adjust decisions in a biological market

### 2022-07-20 manuscript_BE/manuscript_BE.tex
New version of the manuscript

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{float}
\usepackage{booktabs}
\newcommand{\beginsupplement}{ \setcounter{table}{0}     \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}

@@

\setcounter{secnumdepth}{0}

\newcommand{\getAff}[1]{
  \getIndex{#1}{a,b}
}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Field and experimental data together with computational
models reveal how cleaner fish adjust decisions in a biological
market} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Andrés E. Quiñones\textsuperscript{\getAff{a}}\textsuperscript{*},
Zegni Triki\textsuperscript{\getAff{b}},
Redouan Bshary\textsuperscript{\getAff{Institute of Biology, University
of Neuchatel}}\\
\bigskip
\textbf{\getAff{a}}Institute of Biology,University of Neuchatel ,Rue
Emile-Argand 11, 2000-CH Neuchatel, Switzerland\\
\textbf{\getAff{b}}Department of Zoology, Stockholm University, Svante
Arrheniusväg 18 B, 106 91 Stockholm, Sweden,\\
\bigskip
* Corresponding author: andresqp@gmail.com\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
While it is generally straightforward to quantify individual performance
in cognitive experiments, identifying the underlying cognitive processes
remains a major challenge. Often, different mechanistic underpinnings
yield similar performances, and Lloyd Morgan's cannon warrants
acceptance of the simpler explanation. Alternatively, when the different
mechanisms interact with environmental conditions, variation in
performance across environments might allow to statistically infer the
mechanism responsible. We illustrate this point by fitting computational
models to experimental data on performance by wild-caught cleaner fish
\emph{Labroides dimidiatus} in an ephemeral reward task, as well as
cleaner and client fish densities from the locations of capture. Using
Bayesian statistics to fit the model parameters to performance data
revealed that cleaner fish most likely estimate future consequences of
an action, while it appears unlikely that the removal of the ephemeral
reward acts as psychological punishment (negative reinforcement).
Incorporating future consequences also yields performances that can be
considered the result of locally optimal decision-rules, in contrast to
the negative reinforcement mechanism. We argue that the combination of
computational models with data is a powerful tool to infer the
mechanistic underpinnings of cognitive performance.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Performance in behavioural experiments is often used to assess the
cognitive abilities of animals. However, animals can get to the same
outcome in alternative ways. Thus, the outcome of the experiments does
not tell us how animals achieve their performance. In order to overcome
this limitation, we used a set of computational models, which provide
predictions on how alternative cognitive mechanisms perform in the face
of varying environmental conditions, and compare their predictions to
performance data from individuals that experienced different conditions
throughout their life. Our study system is the cleaner fish
\emph{Labroides dimidiatus}, a coral reef fish that feeds off
ectoparasites and dead tissue from the skin of other reef fishes.
Cleaner fish often have to choose among alternative clients seeking
their cleaning service, and the optimal choice is often hard to achieve.
Our performance experiments mimic the natural choices the cleaner fish
make. The combination of experiments and computational models points to
cleaner fish abilities to account for the future effect of their choices
by estimating the long term expected value of those choices. Estimating
long term value is a mechanism involved in human foresight.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
\emph{Text based on plos sample manuscript, see
\url{https://journals.plos.org/ploscompbiol/s/latex}}

@@

Often alternative cognitive mechanisms yield similar behaviour and/or
cognitive performances. This poses a problem for disentangling the
mechanistic underpinnings of behaviour. This is particularly clear in
research aimed at discovering between species variation in \emph{higher}
cognitive abilities; or in other words, research on whether non-human
animals show cognitive abilities believed to be uniquely human. For
instance, when researchers try to find \emph{mental time travel-like}
behaviour, they usually come up with clever experiments to show the
behaviour displayed requires inferences made through past events{[}1{]}.
However, they often face the challenge of alternative scenarios where
simpler explanations, like classic associative learning, can bring about
the observed behavioural outcome {[}2{]}. Similarly, attempts to
demonstrate the presence of \emph{theory of mind} in non-human animals
face objections justified by alternative mechanisms underpinning similar
behavioural results {[}3{]}. Such controversies are usually settled by
using the principle of parsimony and its cognitive version, Lloyd
Morgan's cannon, which states that the simpler explanation (mechanism)
should be accepted. Ideally, however, alternative hypotheses should be
evaluated in light of their explanatory power. One potential dimension
that could allow the variation necessary among distinct cognitive
mechanisms is environmental variation. Controlled laboratory conditions
often mask between and within species variation that arises among
different environments. Thus, if predictions from the alternative
mechanisms depend on environmental conditions, it is possible to assess
their relative explanatory power by performing cognitive test in the
context of different environmental conditions.

From an evolutionary perspective, mechanisms are likely selected because
of how they allow individuals to respond to environmental variation. For
example, biological market theory predicts that the exchange rate of
goods and/or services traded between cooperative partners adjusts to the
law of supply and demand, when individuals have some degree of partner
choice {[}4{]}. Supply and demand conditions, which typically depend on
the abundance of the species involved, certainly vary in time and space.
Therefore, natural selection should favour the ability to flexibly
adjust decisions and behavioural output to current market conditions.
Indeed, such adjustments have been repeatedly documented {[}5{]}. In
animals, an obvious general candidate mechanism for the strategic
adjustment is the cognitive machinery. However, it is not clear which
cognitive mechanisms allow individuals to adjust their behaviour to the
varying conditions. More specifically the question arises to what extent
mechanisms beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides
dimidiatus} and `client' fish. Client fish seek cleaner fish services at
their territory (so-called ``cleaning station'') and offer themselves as
food patches to get their ectoparasites removed, which provides cleaners
with food and clients with improved health {[}6--9{]}. Given the
capacity of some client fish to swim larger distances and access
multiple cleaning stations while others access the only cleaning station
in their territory, it is crucial to categorize clients as either
``visitors'' or ``residents,'' respectively. During cleaning
interactions, a cleaner fish often face a choice between a visitor and a
resident client seeking its cleaning services simultaneously. Visitors
have the option to switch to another cleaner fish if being made to wait,
while residents must wait for inspection. Indeed, visitors have been
observed to use their partner choice option in that way {[}10{]}, which
may explain why cleaners give visitors service priority in a field study
in the Red Sea {[}11{]}.

When researchers aimed at testing wild-caught cleaner fish, as well as
individuals from other species, abilities to prioritize visitor over
resident clients in a lab-based paradigm, they used Plexiglas plates of
different colours and shapes offering the same amount of food as
surrogates for visitor and resident clients. One plate is made to behave
like a resident, i.e.~it remains until the cleaner fish has eaten off
it. The other plate was made to behave like a visitor, i.e.~it is
removed if not inspected first. This laboratory paradigm originally
inspired by the cleaning market of cleaner fish is more generally
referred as the ephemeral reward task. Individuals are assessed in their
capacity to develop a preference for the ephemeral option, which gives
them access to two reward units. Cleaners learned to prefer the visitor
plate and hence obtained the double amount of food {[}12{]}.
Furthermore, adult cleaner fish outperformed various primates as well as
rats and pigeons in this original version of the ephemeral reward task
{[}13,14{]}; yet the African grey parrots solve this task as well
{[}15{]}. This between variation in performance, and primates' known
cognitive abilities {[}\textbf{reader\_Evolution\_2001?}{]}, shows that
performance in the task cannot be readily explained by classifying
species as having high and low cognitive abilities. Instead, this
between species variation suggests that solving the ephemeral reward
task requires a set of cognitive mechanisms that are not present in all
the tested taxa, or were not used under the experimental set-up.

Recent research on the cognitive abilities needed to solve the ephemeral
reward task has used a broad approach as proposed by Shettleworth
{[}16{]}, who defined cognition as including all ways in which animals
acquire information through the senses, process, retain and decide to
act on it. Indeed, perception of relevant cues is of major importance
for individual performance. For example, some animals find relevant
information in the plates {[}17{]} while others find it in the food
{[}18,19{]}. However, identifying cues as salient is not the only
challenge of the ephemeral reward task, as revealed by proximate
learning models. Such models allow varying the cognitive tool kit and
evaluating which minimal kit is necessary to solve the task at hand
(e.g. {[}20{]}). Applied to the ephemeral reward task, learning models
showed that basic reinforcement learning does not suffice to solve the
ephemeral reward task {[}21,22{]}. This is particularly so when models
assume the more complex natural situation in which cleaner fish face not
only resident-visitor pairs but also visitor-visitor and
resident-resident pairs, as well a resident alone or a visitor alone. To
be able to give visitors priority over residents, cleaners need to be
able to assess a client's value separately for the three possible
scenarios (alone, paired with a fish with the same strategic option,
paired with a fish with the alternative strategic option) {[}22{]}. The
ability to distinguish and value one stimulus differently alone from
compound versions of it has been termed configurational learning,
chunking, or segmentation (see references in {[}21{]}).

In addition to configurational learning, cleaners also need to account
for the future consequences of current decisions. In the model by
Quiñones \emph{et al.} {[}22{]}, this could be achieved in two
non-mutually exclusive ways: through low temporal discounting of future
effects, also termed `chaining' {[}23{]}; and/or through perceiving a
visitor client leaving as psychological punishment (i.e.~as a negative
reinforcer). Low temporal discounting is when individuals include in
their valuation of an action the reward effects that this will have in
the future. This is done by combining in a single valuation the reward
obtained in the current time with all the reward that comes after,
discounting for how far in the future reward is accrued. `Chaining' the
reward of these different time steps allows individuals to take actions
that increase the long-term reward at the sacrifice of short term
considerations {[}23{]}. Even though, `chaining' can be readily
implemented computationally in learning models {[}23,24{]}, cognitively
it seems to be a complex adaptation {[}2{]}. On the other hand, using
client behaviour as a negative reinforcer is, in principle, easier to
implement. Negative reinforcement is part of the ubiquitous learning
mechanism termed operant conditioning {[}25,26{]}. Thus, the standard
logic of Lloyd Morgan's cannon demands that operant conditioning as the
simpler explanation is to be accepted by default. Ideally, however, the
two mechanisms should be evaluated in light of how well they explain the
available data. Note that different fields interested in cognition and
decision making use different words to refer to negative reinforcers
{[}22,24{]}. Here, for the sake of simplicity and clarity, we will use
the word `penalty' to refer to this mechanism which includes a negative
reinforcer.

Over the last decade, our team tested over a hundred wild-caught cleaner
fish in the exact same paradigm of the ephemeral reward task
{[}13,27--30{]}. These fish often come from different reef locations
around the study field site at Lizard Island, Great Barrier Reef in
Australia. Further investigation of the local eco-sociological
conditions revealed that cleaner and client fish population densities
have a substantial impact on cleaner fish performance in the task.
Cleaner fish from reef sites with relatively low densities were more
likely to fail at solving the task {[}27--29{]}. The explanation for
this pattern is twofold. First, low cleaner fish population densities
imply low supply and high demand for cleaning services. Under such
conditions, there are fewer occasions under which failing to give
priority to a visitor translates to an empty cleaning station. Second
and also related to the state of the cleaning market, visitor clients at
low cleaner fish population densities are less likely to exert their
partner choice options and are hence more likely to wait for inspection
if not given cleaning priority {[}30{]}. In the absence of partner
choice cleaner fish should not give priority to visitors any more.

The interplay between the underlying cognitive machinery and local
ecological conditions apparently generate the documented variable
performance among individuals of the same species. Our approach of
fitting the computational model to the empirical data on fish densities
and cleaner fish censuses and performance in the ephemeral reward task
aimed at: (i), determining which mechanism cleaner fish use to
incorporate future consequences of current decisions by testing whether
chaining, penalty, or a combination of both best explains their
performance; (ii) determining whether the two mechanisms differed with
respect to the ecological conditions that are likely to cause high
versus low performance in the ephemeral reward task. Additionally, we
assessed which mechanism yields optimal performance patterns. Relying on
the logic of biological market theory, we predicted that appropriate
performance is to show a low preference for visitors under low local
cleaner-to-client ratio.

@@

In this study, our main aim was to unravel which of two potential
cognitive mechanisms, chaining of events, penalty, or their combination,
best explains wild-caught cleaner fish performance in the ephemeral
reward task, while accounting for their ecological conditions. To
evaluate the merits of each of these two mechanisms separately and
combined, we considered cleaner fish performance in the lab test to have
its origin from the rule these fish applied in their natural
environment. That is, individuals that solved the task already had a
preference for visitors clients and generalized this rule to the lab
conditions once being familiar with the task.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{float}
\usepackage{graphicx}
\usepackage{booktabs}
\linespread{2}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}\doublespacing
\newcommand{\beginsupplement}{ \setcounter{table}{0}     \renewcommand{\thetable}{S\arabic{table}}\setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
\providecommand{\keywords}[1]{\textbf{\textit{Keywords---}} #1}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={ Computational models reveal how cleaner fish adjust decisions in a biological market},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\large \textbf{Title:} Computational models reveal how cleaner
fish adjust decisions in a biological market}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\normalsize \textbf{Abbreviated title:} Cleaner fish cognitive
mechanisms in a biological market}
\date{}

@@

\newcommand{\getAff}[1]{
  \getIndex{#1}{}
}

\begin{document}
\maketitle
\begin{flushleft}
\end{flushleft}
\begin{abstract}
While it is generally straightforward to quantify individual performance
in cognitive experiments, identifying the underlying cognitive processes
remains a major challenge. Often, different mechanistic underpinnings
yield similar performances, and Lloyd Morgan's cannon warrants
acceptance of the simpler explanation. Alternatively, when the different
mechanisms interact with environmental conditions, variation in
performance across environments might allow to statistically infer the
mechanism responsible. We illustrate this point by fitting computational
models to experimental data on performance by wild-caught cleaner fish
\emph{Labroides dimidiatus} in an ephemeral reward task, as well as
cleaner and client fish densities from the locations of capture. Using
Bayesian statistics to fit the model parameters to performance data
revealed that cleaner fish most likely estimate future consequences of
an action, while it appears unlikely that the removal of the ephemeral
reward acts as psychological punishment (negative reinforcement).
Incorporating future consequences also yields performances that can be
considered the result of locally optimal decision-rules, in contrast to
the negative reinforcement mechanism. We argue that the combination of
computational models with data is a powerful tool to infer the
mechanistic underpinnings of cognitive performance.
\end{abstract}
\subsection*{Lay summary}
Cleaner fish eat ectoparasites off other fishes, so-called clients. It
regularly happens that two clients seek a cleaner's service
simulatenously. Cleaners benefit from prioritising clients unwilling to
wait, so they can feed on those willing to wait. To make the right
choice, cleaners must somehow ``look'' into the future to anticipate
consequences of current choices. By combining a learning model with
data, we show that cleaners estimate the long-term value of their
actions rather than using simpler heuristics. Estimating long-term value
is a mechanism involved in human foresight.

\keywords{learning, behavior, cleaners, bayesian statisitics, behavioral mechanisms}

@@

Often alternative cognitive mechanisms yield similar behavior and/or
cognitive performances. This poses a problem for disentangling the
mechanistic underpinnings of behavior. This is particularly clear in
research aimed at discovering between species variation in \emph{higher}
cognitive abilities; or in other words, research on whether non-human
animals show cognitive abilities believed to be uniquely human. For
instance, when researchers try to find \emph{mental time travel-like}
behavior, they usually come up with experiments to show the behavior
displayed requires inferences made through past events (Dally, Emery,
and Clayton 2006). However, they often face the challenge of alternative
scenarios where simpler explanations, like classic associative learning,
can bring about the observed behavioral outcome (Suddendorf and
Corballis 2007). Similarly, attempts to demonstrate the presence of
\emph{theory of mind} in non-human animals face objections justified by
alternative mechanisms underpinning similar behavioral results (C. M.
Heyes 1998). Such controversies are usually settled by using the
principle of parsimony and its cognitive version, Lloyd Morgan's cannon,
which states that the simpler explanation (mechanism) should be
accepted. Ideally, alternative hypotheses should be evaluated in light
of their explanatory power.

Learning is a key overarching cognitive mechanism that allows
individuals to associate rewards with environmental stimuli and thus
behave adaptively (Staddon 2016; Shettleworth 2009). Associative
learning, in particular, exists in all major vertebrate taxa, and in
many invertebrates as well (C. Heyes 2012; Macphail 1982; Staddon 2016;
Behrens et al. 2008). Associative learning is not homogeneous throughout
its taxonomic distribution, rather there are differences across and
within species (Shettleworth 2009; Sih and Giudice 2012). So presumably,
the mechanistic underpinnings of learning have been modified by natural
selection (Marler and Peters 1989).

One way to formalize the alternative mechanistic underpinnings of
associative learning is to develop quantitative models of learning
processes. This approach, which started within experimental psychology
(Staddon 2016; Brandon, Vogel, and Wagner 2002), has been very fruitful
in disentangling the mechanistic structure of cognitive systems. More
recently, the development of reinforcement learning theory (Sutton and
Barto 2018), has allowed to evaluate the empirical support for
alternative mechanistic hypotheses by providing quantitative predictions
which are amenable to statistical tests (Farashahi et al. 2020, 2017).
Interestingly, these learning models not only have received support from
behavioral data, but also are consistent with the current view on reward
processing in the brain (Schultz 2015).

From an evolutionary perspective, mechanisms are likely selected because
of how they allow individuals to respond to environmental variation. For
example, biological market theory predicts that the exchange rate of
goods and/or services traded between cooperative partners adjusts to the
law of supply and demand, when individuals have some degree of partner
choice (Noë and Hammerstein 1995). Supply and demand conditions, which
typically depend on the abundance of the species involved, certainly
vary in time and space. Therefore, natural selection should favor the
ability to flexibly adjust decisions and behavioral output to current
market conditions. Indeed, such adjustments have been documented (Axén,
Leimar, and Hoffman 1996). One example of strategic adjustment in a
biological market is the marine cleaning mutualism involving the cleaner
fish \emph{Labroides dimidiatus} and `client' fishes. Client fishes seek
cleaner fish services at their territory (so-called ``cleaning
station'') and offer themselves as food patches to get their
ectoparasites removed, which provides cleaners with food and clients
with improved health (Waldie et al. 2011; Ros et al. 2011; Triki et al.
2016; Demairé et al. 2020). Given the capacity of some client fish to
swim larger distances and access multiple cleaning stations while others
access the only cleaning station in their territory, it is crucial to
categorize clients as either ``visitors'' or ``residents'',
respectively. During cleaning interactions, a cleaner fish often faces a
choice between a visitor and a resident client seeking its cleaning
services simultaneously. Visitors have the option to switch to another
cleaner fish if being made to wait, while residents must wait for
inspection. Indeed, visitors have been observed to use their partner
choice option in that way (Bshary and Schäffer 2002), which may explain
why cleaners give visitors service priority in a field study in the Red
Sea (Bshary 2001). Furthermore, in a lab based paradigm, design to mimic
the resident-visitor choice (ephemeral reward task), cleaners learned to
prefer the cue associated with the epheral food source (visitor) and
hence accessed both food sources, obtaining double the amount of food
(Bshary and Grutter 2002). However, further exploration revealed that
not all cleaner fish manage to develop a preference for the ephemeral
option in the lab (Triki et al. 2018). Over the last decade, over a
hundred wild-caught cleaner fish have been tested in the exact same
paradigm of the ephemeral reward task (Salwiczek et al. 2012; Wismer et
al. 2014; Triki et al. 2018, 2019, 2020). These fish often come from
different reef locations. Investigation of the local eco-sociological
conditions revealed that cleaner and client fish population densities
have a substantial impact on cleaner fish performance in the task.
Cleaner fish from reef sites with relatively low densities were more
likely to fail at solving the task (Triki et al. 2019, 2018; Wismer et
al. 2014). This intra-specific variation is unlikely due to local
genetic adaptation, because cleaner fish are open water spawners and the
environmental conditions can vary within the lifespan of a fish.

Mechanistic models explicitly designed to mimic the ephemeral reward
task have shown that the simplest form of associative learning (operant
conditioning) cannot account for a solution to the ephemeral reward task
(Prat, Bshary, and Lotem 2022; Quiñones et al. 2019). Operant
conditioning is a form of associative learning where individuals use
short term reward to associate and choose actions. Such models allow
varying the cognitive tool kit and evaluating which minimal kit is
necessary to solve the task at hand (e.g. Dubois et al. (2021)). To be
able to give visitors priority over residents, cleaners need to be able
to assess a client's value separately for the three possible scenarios
(alone, paired with a fish with the same strategic option, paired with a
fish with the alternative strategic option) (Quiñones et al. 2019). The
ability to distinguish and value one stimulus differently alone from
compound versions of it has been termed configurational learning,
chunking, or segmentation (see references in Prat, Bshary, and Lotem
(2022)). In addition to configurational learning, cleaners also need to
account for the future consequences of current decisions. In the model
by Quiñones et al. (2019), this could be achieved in two non-mutually
exclusive ways: through low temporal discounting of future effects, also
termed `chaining' (Enquist, Lind, and Ghirlanda 2016); and/or through
perceiving a visitor client leaving as psychological punishment (i.e.~as
a negative reinforcer). Chaining is when individuals include in their
valuation of an action the reward effects that this will have in the
future. This is done by combining in a single valuation the reward
obtained in the current time with all the reward that comes after,
discounting for how far in the future reward is accrued. `Chaining' the
reward of these different time steps allows individuals to take actions
that increase the long-term reward at the sacrifice of short term
considerations (Enquist, Lind, and Ghirlanda 2016). Even though,
`chaining' can be readily implemented computationally in learning models
(Enquist, Lind, and Ghirlanda 2016; Sutton and Barto 2018), cognitively
it seems to be a complex adaptation (Suddendorf and Corballis 2007). On
the other hand, using client behavior as a negative reinforcer is, in
principle, easier to implement. Thus, the standard logic of Lloyd
Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Ideally, however, the two
mechanisms should be evaluated in light of how well they explain the
available data. Note that different fields interested in cognition and
decision making use different words to refer to negative reinforcers
(Quiñones et al. 2019; Sutton and Barto 2018). Here, for the sake of
simplicity and clarity, we will use the word `penalty' to refer to this
mechanism which includes a negative reinforcer.

In here we used the field and experimental data to fit the parameters of
a reinforcement learning model to infer the cognitive mechanism that
cleaners use in their interaction with clients. Specifically, our
approach of fitting the computational model to the empirical data aimed
at: (i), determining which mechanism cleaner fish use to incorporate
future consequences of current decisions by testing whether chaining,
penalty, or a combination of both best explains their performance; (ii)
determining whether the two mechanisms differed with respect to the
ecological conditions that are likely to cause high versus low
performance in the ephemeral reward task. Additionally, we assessed
which mechanism yields optimal performance patterns. Relying on the
logic of biological market theory, we predicted that appropriate
performance is to show a high preference for visitors only under high
local cleaner-to-client ratio.

@@

In this study, our main aim was to unravel which of two potential
cognitive mechanisms, chaining of events, penalty, or their combination,
best explains wild-caught cleaner fish performance in the ephemeral
reward task, while accounting for their ecological conditions. To
evaluate the merits of each of these two mechanisms separately and
combined, we considered cleaner fish performance in the lab test to have
its origin from the rule these fish applied in their natural
environment. That is, individuals that solved the task already had a
preference for visitor clients and generalized this rule to the lab
conditions once being familiar with the task.
